This is the first Peer Reviewed Assignment for the Reproducible Research Course
===============================================================================
First, set the working directory, prepare libraries and read in the data:
```{r}
library(dplyr)
library(tidyr)
library(reshape2)
setwd("c:/RR/ReproRes/repdata_data_activity")
activity <- read.csv("activity.csv",header=TRUE)
```
## Creating dataset with missing values removed
```{r}
step_data <- activity[complete.cases(activity),]
step_data$interval <- sprintf("%04d", step_data$interval)
step_data$int_time <- strptime(step_data$interval, format = "%H%M")

```
## Histogram of step count per day, missing values removed
```{r}
step_data <- transform(step_data, dayf=factor(date))
daysum <- tapply(step_data$step, step_data$dayf, sum)
hist(daysum, xlab = "Step count per day")
```
## Step up data frame to calculate Mena and Median steps per day:

# First calculate total no of steps per day
```{r}
step_sum <- tapply(step_data$steps,step_data$dayf, sum)
daily_steps <- as.data.frame(step_sum)
```
# Next calculate median of total step count per day
```{r}
step_median <- quantile(daily_steps$step_sum,probs=0.5)
print("Median steps per day = ")
print(step_median)
```
# And mean of total step count per day:
```{r}
step_mean <- mean(daily_steps$step_sum)
print("Mean steps per day = ")
print(step_mean)
```
## Average daily activity pattern
# Plot of average number of steps in each 5 minute interval
```{r}
step_data <- transform(step_data, intervalf=factor(interval))
intmean <- tapply(step_data$step, step_data$intervalf, mean)
plot(intmean, type="l", xlab="5 minute interval", ylab = "Average number of steps")
```
# Interval where average number of steps is maximum
```{r}
maxstep = max(intmean)
max_index <- which(intmean == max(intmean))
max_interval <- step_data$interval_revised[max_index]
max_interval
```
## Imputing missing data
# First, get count of missing data
```{r}
total_length <- dim(activity)[1]
present_length <- dim(step_data)[1]
missing_count <- total_length - present_length
print("Missing = ")
missing_count
```
# Replacing the missing values with the 5 minute interval means
# Make wide version of actitity data frame
```{r}
widestep <- dcast(activity, date ~ interval, value.var = "steps")
wsdate <- widestep$date
names(widestep) <- paste("time", names(widestep), sep = "")
for (i in which(sapply(widestep, is.numeric))) {
  widestep[is.na(widestep[, i]), i] <- mean(widestep[, i],  na.rm = TRUE)
}
widestep$date <- wsdate
```
# Make long version of activity with missing replace by mean for interval
```{r}
step_nomiss <- widestep %>%
  gather(interval, step, time0:time2355)
```
## Histogram of step count per day for data with missing values replaced by interval means
```{r}
step_nomiss$interval <- as.numeric(substr(step_nomiss$interval, 5,8))
step_nomiss <- transform(step_nomiss, dayf=factor(date))
daysum <- tapply(step_nomiss$step, step_nomiss$date, sum)
hist(daysum, xlab = "Step count per day imputted missing")
```
# Mean and Median steps per day for dataset with missing values
# replaced by interval mean:
## First calculate total no of steps per day
```{r}
step_summi <- tapply(step_nomiss$step,step_nomiss$dayf, sum)
daily_stepsmi <- as.data.frame(step_summi)
```
# Calculate median and mean for total daily steps, imputted missing
```{r}
step_median_mi <- quantile(daily_stepsmi$step_summi,probs=0.5)
print("Median steps per day = ")
print(step_median_mi)
print("Mean steps per day = ")
step_mean_mi <- mean(daily_stepsmi$step_summi)
print(step_mean_mi)
```
# Confirm that replacing missing data has had no effect on means and little effect on the median
```{r}
meandiff <- abs(step_mean_mi - step_meane)
meddiff <- abs(step_median_mi - step_mediane)
meandiff
meddiff
```
## Investgating the difference in activity pattern between weekdays
## and weekends

# create factor for differentiating between weekdays and weekend days
```{r}
step_nomiss$timedate <- as.Date(step_nomiss$timedate,format="%Y-%m-%d")
step_nomiss$day <- weekdays(step_nomiss$timedate)
step_nomiss$weekday <- (step_nomiss$day == "Saturday" | step_nomiss$day== "Sunday")
step_nomiss$weekday <- as.factor(step_nomiss$weekday)

step_nomiss$intday <- as.factor(paste(step_nomiss$weekday,step_nomiss$interval))
```
## Calculate mean 5 minute interval steps counts for weekends and weekdays and plot results
```{r}
suppressWarnings(intmean2 <- aggregate(step_nomiss, by = list(step_nomiss$interval, step_nomiss$weekday), mean))
intmean2$day <- rep(c("Weekend", "Weekday"), each=288)

xyplot(step ~ interval | day, data = intmean2, layout = c(1,2), type="l",ylab="Number of steps")
```